---
title: 'Community Ecology Part 1: Alpha diversity of a Jellybelly Community'
author: "Kristen DeAngelis"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document: default
editor_options:
  chunk_output_type: console
---

## Overview

This is a lab activity that will read in data from a single jellybelly community, use a variety of methods to explore the data, and then calculate some popular measures of alpha diversity for this single sample. 

Some of this code was adapted from "Jellybelly Diversity" by "Jay T. Lennon" (https://github.com/LennonLab/jellybelly) and (https://github.com/QuantitativeBiodiversity/QB-2017/tree/master/Week2-Alpha). 


# Directions:
1. Change "author" on line 3 (above) to your name.
2. Complete as much of the assignment as possible during class; what you do not complete in class will need to be done on your own outside of class. 
3. Use this handout as a guide; it contains a more complete description of data sets along with examples of proper scripting needed to carry out the exercise.
4. Be sure to **answer the questions** in this assignment document. Space for your answers is provided in this document and is indicated by the ">" character. If you need a second paragraph be sure to start the first line with ">".

As we begin this module, you should have already completed the "Measuring Diversity Activity: the Jellybean Community." You should have also entered your data into the class spreadsheet, which we will read in to analyze our data. 

We are working in Rstudio, and will work using markdown. R markdown files are denoted by the file extension .Rmd. Markdown is a simple but powerful formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on how to get the most out of the Markdown language in R, see <http://rmarkdown.rstudio.com>.

In this Handout, we will perform the following tasks:

 1. Set up our environment and load the libraries we'll need. 
 2. Read in files. 
 3. Explore new object contents and structure.
 4. Richness calculations
 5. 

First we will set up our environment and load the libraries we'll need. 

```{r setup, results = 'hide'}
# rm(list = ls())

```

## 2. Load data

This first chunk will create a model dataset for a mock jellybelly community. 

```{r create a model dataset}
flavors <- c("Berry Blue", "Blueberry", "Bubble Gum", "Buttered Popcorn", "Cantaloupe", "Cappuccino", "Caramel Corn", "Chocolate Pudding", "Cinnamon", "Coconut", "Cotton Candy", "A&W Cream Soda", "Crushed Pineapple", "Dr Pepper", "French Vanilla", "Green Apple", "Island Punch", "Juicy Pear", "Kiwi", "Sunkist Lemon", "Lemon Drop", "Lemon Lime", "Licorice", "Sunkist Lime",  "Mango", "Margarita", "Mixed Berry Smoothie", "Sunkist Orange", "Orange Sherbet", "Peach", "Pina Colada", "Sunkist Pink Grapefruit", "Plum", "Pomegranate", "Raspberry", "Red Apple", "A&W Root Beer", "Sizzling Cinnamon", "Sour Cherry", "Strawberry Cheesecake", "Strawberry Daquiri", "Strawberry Jam", "Sunkist Tangerine", "Toasted Marshmallow", "Top Banana", "Tutti-Fruitti", "Very Cherry", "Watermelon", "Wild Blackberry")

set.seed(123)  # for reproducibility
jellybean_counts <- sample(10:100, size = length(flavors), replace = TRUE)
names(jellybean_counts) <- flavors

```

When you sampled your jellybean community, you should have recorded morphotypes (or species) as well as number of samples in each category. However, we're going to use a conventional species defintion for species here, so you should have gone back through and id'ed your individuals based on your best guess using the Jelly Belly flavor guide: https://www.jellybelly.com/flavor-guides

In each cell of the matrix, the number denotes the count per species, which is the number of beans in a sample belonging to morphotype. All the data should be loaded into the spreadsheet at the beginning of lab. We will read in the matrix of data for the whole class, then subset the data to include only the vector of data containing your sampling. 

```{r }
jb.raw <- read.table("jellybelly.txt", sep = "\t", header = TRUE)
```


## 3. Explore data 

Before we move on, we will explore the data. For a small dataset like this, it is easy to look it over by eye. But when doing community analysis for complex communities like microbes, it is common to have data matrixes with thousands of rows. 

Explore the data. This is a small dataset, but in microbial ecology, it is usual to have thousands to tens of thousands of taxa. These datasets are much to large to view directly, so they have to be explored. 

Think of data exploration as a set of hypotheses that you test, in an iterative progression, with each experiment including expected results and evidence for whether you can accept or reject each hypothesis. 

This is a function in base R that reports the object kind or class. We created this dataset so it should be class "integer."
```{r}
class(jellybean_counts)
x <- jellybean_counts
```
What kind of object is my data? 
> 

This is a function in base R that reports the length of a vector or any other R object for which lenght has been defined. 
```{r}
length(x)        # Number of elements
head(x)          # First few values
tail(x)          # Last few values
summary(x)       # Min, 1st Qu., Median, Mean, 3rd Qu., Max
```
What is the length of my data? 
>

These are some descriptive statistics that can give a sense of the central tendencies of the data. 
```{r}
mean(x)          # Average
median(x)        # Middle value
sd(x)            # Standard deviation
var(x)           # Variance
range(x)         # Min and max
quantile(x)      # Percentiles
```


These are some visualizations that can be used to see impressions of the data, which are particularly helpful for very large datasets. 
```{r}
hist(x)                  # Histogram
boxplot(x)               # Boxplot
plot(density(x))         # Density plot
barplot(table(x))        # Barplot for categorical or rounded data


```


What are the variable names? What is the structure of my data object? 


## 2) DIVERSITY CALCULATIONS

```{r}
# Total abundance (N)
sum(jb$count)

# Observed richness
S.obs <- function(x = ""){
  rowSums(x > 0) * 1
}

S.obs(t(jb[,5]))

# Coverage
coverage <- function(x = ""){
  1 - (sum(x == 1) / rowSums(x))
  }

coverage(t(jb$count)) 

# Esimated richness with Chao1
S.chao1 <- function(x = ""){
  S.obs(x) + (sum(x == 1)^2) / (2 * sum(x == 2))
}

S.chao1(t(jb[,5]))

# Visualize rank abundance curve
RAC <- function(x = ""){
  x = as.vector(x)
  x.ab = x[x > 0]
  x.ab.ranked = x.ab[order(x.ab, decreasing = TRUE)]
  return(x.ab.ranked)
  }

plot.new()
site1 <- t(jb[,5])

rac <- RAC(x = site1)
ranks <- as.vector(seq(1, length(rac)))
opar <- par(no.readonly = TRUE)                       # Saves default plot parameters
par(mar = c(5.1, 5.1, 4.1, 2.1))                      # New settings for par
plot(ranks, log(rac), type = 'p', axes = F,           # Plots w/o axes
     xlab = "Rank in abundance", ylab = "Abundance",
     las = 1, cex.lab = 1.4, cex.axis = 1.25)

box()                                                 # Manually adds border
axis(side = 1, labels = T, cex.axis = 1.25)           # Manually adds X-axis
axis(side = 2, las = 1, cex.axis = 1.25,              # Manually adds Log-Scaled Y-axis
     labels = c(1, 2, 5, 10, 20), at = log(c(1, 2, 5, 10, 20)))

par <- opar                                           # Resets plotting parameters

# Smith and Wilson's Evenness Index (Evar)
Evar <- function(x){
  x <- as.vector(x[x > 0])
  1 - (2/pi)*atan(var(log(x)))
  }

Evar(t(jb[,5]))

# Shannon's Diversity (H')
ShanH <- function(x = ""){
  H = 0
  for (n_i in x){
    if(n_i > 0) {
      p = n_i / sum(x)
      H = H - p*log(p) 
    }
  }
  return(H)
}

ShanH(t(jb[,5]))
```


